Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-12-18 demystify.tex
updates

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{soul}
% \usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
% \setsecnumdepth{subsection}
% \setcounter{tocdepth}{3}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{pdfpages}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float

@@

\newpage
\tableofcontents

% \chapter{Summary}\label{sec:summary}

% This section summarizes the main points of this guideline. It also gives you an overview to decide which specific topics you want to explore more thoroughly.

  %\item Should you apply to a PhD CS program in the US?

          %\item \emph{Contacting a prof. is recommended}, but do it \emph{properly} to get a reply (\autoref{sec:contact}).

          %\item  (\autoref{sec:interview}).

    % \item Miscs and FAQS
    %     \begin{itemize}

    %       \item (\autoref{sec:read-all}).

    %       \item
    %       \item
    %       \item (\autoref{sec:msrequirement}), and CS PhD takes abou (\autoref{sec:time}). Pu(\autoref{sec:non-us-differences}).
    %       \item
    %       \item
    %       \item
    %       %\item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
    %     \end{itemize}

    % \item Appendices
    % \begin{itemize}
    %   \item Specific advice for domestic students (\autoref{sec:domestic-students}).
    %   \item How to find research opportunities (\autoref{sec:research-opportunities}).
    %   \item Cultural differences between US and other countries (\autoref{sec:cultural}).
    % \end{itemize}

@@

\part{Introduction To CS PhD Admission in the US}\label{part:intro}

@@

\chapterinfo{Yes, definitely.  CS PhD study in the US is fully funded and admission into good universities in the US is not any harder than in other countries.}

@@

\chapterinfo{Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members.}

@@

\section{Admission Committee (adcom)}\label{sec:adcom}
\sectioninfo{Adcom members are faculty who evaluate your application.}

Your applications are reviewed by a PhD admission committee or \emph{adcom} that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At Mason, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review about 30 applications. Note that most large schools, including Mason, have separate adcoms for MS programs (\autoref{chap:ms}).

Each application is assigned to about three adcom members, who will evaluate your profile and reach a consensus.  While the assigned reviewers are the main ones deciding your application, other faculty in the department can also have access to your application and provide inputs and opinions on your profile.

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various types of faculty). This provides junior faculty the opportunities to recruit students. The chair of the committee will likely be a senior faculty, but they will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty, e.g., I am often assigned to review applicants interested in software engineering.

\begin{commentbox}{Vu:}
At Mason, we usually decide that a full-time PhD candidate is either (i) admitted with funding (\autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding or reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision (\autoref{sec:ievaluate}) with a summary of your application, where we list
strengths, e.g., a well-known school, and weaknesses, e.g., weak
LORs.
\end{commentbox}

@@

@@

\section{LoR writers}\label{sec:lor-writers}

@@

\section{Asking for LORs}\label{sec:asking-lor}

@@

Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English. 
You might be asked to talk about your research experience and interests and to read a paper and discuss it. In some rare cases you might also be asked to solve a problem (one of my colleagues at Mason often does coding interview).
The interview is also a good opportunity for you to ask questions, e.g., about the prof, their group, the CS PhD program, and the university.

You should treat the interview as an informal chat. Have ``elevator pitches'' about your research experience and interests. You might also want to have a 5-minute presentation about your research. If a prof. asks you to read a paper, do it and be prepared to discuss it. You can also ask if you need to prepare for a coding problem. Finally, prepare some thoughtful questions to ask, e.g., about the program, the university, or the professor's research (see \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Answers-to-Ph.D-Advisor-Guide}{some questions} you can ask a potential adviser).

@@

@@

\section{If you do not get admitted}\label{sec:not-accepted}

@@

\subsection{Why did you get rejected?}\label{sec:why-rejected}
\subsectioninfo{You aim too high, are overqualified, or even because you applied to AI/ML, a super competitive field in recent years with many applicants.}

@@

\paragraph{You aim too high} 
You have applied to schools that are \emph{way too high} for your profile (\autoref{sec:selecting-ranking-schools}). Many students simply just go for very top schools (e.g., top 10) and are surprised when they are rejected to all of them, in multiple years, and completely get shut out.  This is very obvious but many students still do this. An analogy is a person who has never hike wanting to climb Everest, which btw if you could, you might have a better admission chance (\autoref{sec:improve-your-chance}).

@@

\subsection{Increasing your admission chance}\label{sec:improve-your-chance}
\subsectioninfo{You can improve your profile by being unique and standing out.}

@@

\part{Funding, Schools, and Profs.}\label{part:funding-schools-profs}

>>>>

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{soul}
\usepackage[most]{tcolorbox}
% \setsecnumdepth{subsection}
% \setcounter{tocdepth}{3}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{pdfpages}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float

@@

\newpage
\renewcommand{\contentsname}{Contents and Summary}
\tableofcontents

@@

\part{Introduction To CS PhD Admission}\label{part:intro}

@@

\chapterinfo{Yes, definitely.  CS PhD study in the US is often fully funded and admission into good universities in the US is not any harder than in other countries.}

@@

\chapterinfo{Applications are evaluated by the \emph{PhD Admission} (\textbf{adcom}) committee and each application is typically reviewed by three faculty members.}

@@

\section{Admission Committee (adcom)}\label{sec:adcom}
\sectioninfo{Adcom members are faculty who evaluate your application. They consider various factors, e.g., research experience, LoRs, SOP}

Your applications are reviewed by a PhD admission committee or \emph{adcom} that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. Adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At Mason, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review about 30 applications. Note that most large schools, including Mason, have separate adcoms for MS programs (\autoref{chap:ms}).

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various types of faculty). This provides junior faculty the opportunities to recruit students. The adcom \emph{chair} will likely be a senior faculty, but they will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty, e.g., I am often assigned to review applicants interested in software engineering.

Each application is assigned to about three adcom members, who will evaluate your profile and reach a consensus.  They will consider various factors, e.g., research experience, GPA, LoRs, SOP, and test scores. More details about these factors are discussed in \autoref{part:appendices}.

\begin{commentbox}{Vu:}
At Mason, we usually decide that a full-time PhD candidate is either (i) admitted with funding (\autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding or reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision (\autoref{sec:ievaluate}) with a summary of your application, where we list
strengths, e.g., a well-known school, and weaknesses, e.g., weak LoRs.
\end{commentbox}

@@

Note that while the assigned reviewers are the main ones deciding your application, other faculty in the department can also have access to your application and provide inputs and opinions on your profile. Thus, it helps to contact faculty (\autoref{sec:contact}) and mention faculty you're interested in in your SOP (\autoref{chap:sop}).

@@

\section{LoR writers}\label{sec:lor-writers} \sectioninfo{LoR writers should be someone who (i) can talk in depth about your research experience and potential and (ii) have the credibility to evaluate your research ability.}

@@

\section{Asking for LORs}\label{sec:asking-lor}
\sectioninfo{Ask for LoRs at least a month before the deadline.  Waive your right to see the letter. Help your writers by providing details about your research experience and potential.}

@@

@@

\paragraph{Preparing for interviews} Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English. 
You might be asked to talk about your research experience and interests and to read a paper and discuss it. In some rare cases you might also be asked to solve a problem (one of my colleagues at Mason often does coding interview).

You should treat the interview as an informal chat. Prepare an \emph{``elevator pitche''} about your research experience and interests. You might also want to have a 5-minute presentation about your research. If a prof. asks you to read a paper, do it and be prepared to discuss it. You should also ask if you need to prepare for coding. Finally, the interview gives you an opportunity to ask questions, e.g., about the program and the professor's research.  You should definitely ask as it shows that you're interested (see \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Answers-to-Ph.D-Advisor-Guide}{some questions} you can ask).

@@

\chapter{Dealing with Rejection}\label{chap:not-accepted}

\section{Try Again!}\label{sec:try-again}

@@

\section{Why did you get rejected?}\label{sec:why-rejected}
\subsectioninfo{You aim too high, are overqualified, or even because you applied to AI/ML, a super competitive field in recent years with many applicants.}

@@

\paragraph{You aim too high} 
You have applied to schools that are \emph{way too high} for your profile (\autoref{sec:selecting-ranking-schools}). Many students simply just go for very top schools (e.g., top 10) and are surprised when they are rejected to all of them, in multiple years, and completely get shut out.  This is very obvious but many students still do this (\autoref{sec:chance-me}). An analogy is a person who has never hike but wants to climb Mt. Everest (which btw if you could, you might have a better admission chance as mentioned in \autoref{sec:improve-your-chance}).

@@

\section{Increasing your admission chance}\label{sec:improve-your-chance}
\subsectioninfo{You can improve your profile by being unique and standing out.}

@@

\part{Types of Funding and Choosing Schools and Profs.}\label{part:funding-schools-profs}
